Typestate Inference in an Imperative Calculus

### 2012-11-27 formalism/paper.tex
proof of context normalisation almost complete, interim commit before final pass

\newcommand{\typerule}[4]{#1 \triangleright #2 : #3 \triangleleft #4}
\newcommand{\oprule}[4]{#1 \mid #2\;\longrightarrow\;#3 \mid #4}
\newcommand{\inferrule}[4]{\left\llbracket #1 \triangleright #2 : #3 \triangleleft #4 \right\rrbracket}
\newcommand{\inferlhs}[1]{\left\llbracket \gamma \triangleright #1 : \alpha \triangleleft \gamma' \right\rrbracket}
\newcommand{\subst}[3]{#3 \{\overline{^{#1}/_{#2}}\}}
\newcommand{\substxy}[1]{\subst{x_i}{y_i}{#1}}
\newcommand{\unitv}{\mathtt{unit}}
\newcommand{\unitt}{\mathbf{Unit}}
\newcommand{\truev}{\mathtt{true}}
\newcommand{\falsev}{\mathtt{false}}
\newcommand{\boolt}{\mathbf{Bool}}
\newcommand{\ot}[2]{#1@\overline{#2}}
\newcommand{\funv}[4]{\lambda(\overline{#1 : #2 \gg #3}).#4}
\newcommand{\funt}[3]{(\overline{#1 \gg #2}) \rightarrow #3}
\newcommand{\lett}[3]{\mathtt{let}\:#1\:\mathtt{=}\:#2\:\mathtt{in}\:#3}
\newcommand{\ift}[3]{\mathtt{if} \; #1 \; \mathtt{then} \; #2 \; \mathtt{else} \; #3}
\newcommand{\cand}{\:\wedge\:}
\newcommand{\free}[1]{\mathbf{free}\:#1}
\newcommand{\fresh}[1]{\mathbf{fresh}\:#1}

@@

% unit value
\inferlhs{\unitv} \triangleq \alpha = \unitt \cand \gamma = \gamma' 
\\\\

% true value
\inferlhs{\truev} \triangleq \alpha = \boolt \cand \gamma = \gamma' 
\\\\

% false value
\inferlhs{\falsev} \triangleq \alpha = \boolt \cand \gamma = \gamma' 
\\\\

@@

% let binding
\inferlhs{\lett{x}{t'}{t}} \triangleq \\
\exists \gamma_v, \gamma_{b1}, \gamma_{b2}, \alpha' . \\
\gamma_{b1} = \gamma_v, x : \alpha'
\cand
\gamma' = \gamma_{b2} - \{ x \}
\cand \\
\inferrule{\gamma}{t'}{\alpha'}{\gamma_v}
\cand
\inferrule{\gamma_{b1}}{t}{\alpha}{\gamma_{b2}}
\\\\

@@

Additionally, we have {\bf variable usage constraints} of form 
$x : \mathbb{T} \in \gamma$, which are generated for each location in the
program where a variable is used and what type that usage implies. These
are used to generate additional type variable
constraints by matching the asserted type $\mathbb{T}$ with what is known
about the type of $x$ in $\gamma$.

@@

Type variables can appear at any position within a type's structure, as shown
in the grammar for $\mathbb{T}$. Objects are treated as a special case, where
we use variables as placeholders for the object definition and state parts.
The states and transitions of the object structure are determined at a later
stage in the type inference process.

@@

\begin{itemize}
\item All value terms produce a $\gamma = \gamma'$ constraint as values do
not change the context. Additionally, a type variable constraint of form
$\alpha = \mathbb{T}$ where $\mathbb{T}$ is of the appropriate type to match
the value, with type variables in all positions relating to parameters and return
values.
\item The context variable $\gamma_a$ used as the input context for
function bodies has an empty root $\gamma_r$, to allow for detection of 
used unbound variables within
the body. This matches the use of a context in type inference
composed only of the parameters, and prevents implicit binding.
\item Let-bind terms generates a new context variable $\gamma_v$ for the
output context of the value term. This is extended by $\gamma_{b1}$, the
input context for the body term, with variable $x$ of the type inferred
for the value. The output context of the whole let-bind term removes
$x$ from the output context $\gamma_{b2}$ of the body term.
\item Update terms generate replacement constraints based on the inferred
type $\alpha'$ of the right-hand side of the expression.
\item Sequencing terms relate the output of the left term to the input of the
right term through a new context variable $\gamma''$. The type variable
$\alpha'$ must be generated to capture the type of the left term for
consistency, but it is not needed.
\item Function call terms generate context variable constraints to relate the
usage of the variables to whatever information has been inferred about the
parameter types of the function. All variables $\overline{x_i}$ used as 
parameters in the function call are replaced in the output context with
new type bindings to $\overline{\alpha_i'}$, which represents the effect
of the function.
\item Method call terms generate type variables to capture that $x$ must
be an object, that the method $m$ must exist in the current state $\alpha_s$
and that after the method call $x$ must be in the new state $\alpha_s'$.
\end{itemize}

@@

Constraint solving is performed in phases. In the first phase, the context 
constraints
are normalised using the rules in \figref{fig:contextsolve} --- where the set of 
constraints in the premise exists, they are replaced with those in the 
conclusion. Normalisation means that every context variable will have a single 
defining constraint of form $\gamma = \emptyset$, $\free{\gamma}$ or
$\gamma = \gamma', \overline{x_i : \mathbb{T}_i}$, where $\gamma'$ is
either empty or free. A context $\Gamma$ for type checking can be easily
constructed from constraints in this form, as once the type variable
constraints are solved we can say $\Gamma = \overline{x_i : T_i}$ for
every $\gamma = \gamma', \overline{x_i : \mathbb{T}_i}$ where $T_i$ is the
solved form of $\mathbb{T}_i$.

@@

\begin{itemize}
\item \textsc{ceq} --- Where $\gamma$ is directly equivalent
to $\gamma'$, which is defined in terms of extension of
$\gamma''$, rewrite the constraints such that
$\gamma$ directly copies the extension of $\gamma''$.
\item \textsc{crem} --- Where $\gamma$ removes a variable $x$ from $\gamma'$,
which is defined in terms of extension of $\gamma''$, rewrite the
constraints such that $\gamma$ copies the extension of $\gamma''$ with
the variable $x$ removed from the extension. There is no need for an
error-producing variant of \textsc{crem}, where the variable $x$ is missing, 
as context removal constraints are always generated in a pair with a context 
extension constraint for let binding.
\item \textsc{cext} --- Where $\gamma$ extends $\gamma'$ with
a variable set $\overline{y_j}$, and $\gamma'$ is defined in terms of extension of
$\gamma''$ with a disjoint set of variables in its extension set, rewrite the
constraints such that $\gamma$'s extension set includes both the $y_j$
definitions and $x_j$ definitions.
\item $\textsc{cext}_2$ --- Where $\gamma$ extends $\gamma'$ with
a new variable $x$, and $\gamma'$ is defined in terms of extension of
some $\gamma''$ and the extension set already contains $x$, this
represents an attempt to redeclare $x$. As such, generate an error
$\mathbf{bound}\:x$, terminating the constraint solving process.
\item \textsc{crep} --- where $\gamma$ replaces the type of some
variable $x$ in $\gamma'$, where $\gamma$ is defined in terms of extension
of $\gamma''$ and $x$ is in the extension set, rewrite the constraints
such that $\gamma$ copies the extension set of $\gamma'$ with $x$
replaced as specified.
\item $\textsc{crep}_2$ --- where $\gamma$ replaces variable $x$ in
$\gamma'$ which is defined in terms of extension of $\gamma''$
which is free and $x$ is not in the extension set, $x$ is a
free variable in the term being processed. To introduce $x$ in such a way that
it will be included in all contexts, $\gamma''$ is redefined in
terms of extension of a new context $\gamma_r$ with variable $x$
in the extension set mapped to a new type variable $\alpha$. Following
this, \textsc{cext} can be applied to $\gamma'$, then \textsc{crep} can
be applied to $\gamma$.
\item $\textsc{crep}_3$ --- where $\gamma$ replaces variable $x$ in
$\gamma'$ which is defined in terms of an empty root $\gamma''$,
and $x$ is not in $\gamma$'s extension set, this indicates the usage
of an unbound variable within a function body. As such, generate an
error $\mathbf{unbound}\:x$ indicating that x is unbound within this
context, terminating the contraint solving process.
\item \textsc{cvar} --- where a variable constraint for $x$ exists
for context variable $\gamma$, which is defined in terms of
extension of $\gamma_r$ which is free, and $x$ is not in the extension
set of $\gamma$, $x$ is a free variable in the term being processed. 
Similar to $\textsc{crep}_2$, redefine $\gamma_r$
in terms of extension of a new root $\gamma_r'$, with $x$ in
the extension set with the observed type $\mathbb{T}$.
\item $\textsc{cvar}_2$ --- where a variable constraint for $x$ exists
for context variable $\gamma$, which is defined in terms of $\gamma'$
which is empty, and $x$ is not in the extension set of $\gamma$, 
$x$ is an unbound variable within a function body.
\end{itemize}

@@

\subsection{Variable usage \& type variable constraint solving}

In the second phase, additional type variable 
constraints are generated by following the rule in \figref{fig:tvargen}, which
compares the observed type from the usage of a variable to the type
known in the context. Repeated application of this rule until no more matches
are found produces a constraint set composed exclusively of normalised
context variable constraints, type variable constraints and protocol
constraints. A first order unification algorithm can then be used to solve
the system of type variable constraints, producing a most general unifier
for the type variables if unification is possible.

First order unification will identify a set of type variables (and their synonyms)
which represent distinct object types. For each distinct object type, there
will also be a set of type variables which represent distinct states for
the object type. By matching these variables to the protocol constraints
within the constraint set, we can generate the state machine for each
object type in a straightforward manner. To produce a minimal description of 
the object protocol, this state machine can be minimised by constructing a 
bisimulation relation between the states within each object and collapsing 
equivalent states.

The typing rules of TS0 do not permit polymorphism, so
if the most general unifier does not map all type variables for
variables that occur in the input and output contexts of the term being
processed, then the term has no valid type in the language. Otherwise,
substituting the fully explicit type inferred for each type variable that
occurs in the inferred input and output context variables, as well as the
type variable for the input term, should produce a valid typing judgement
where the domain of the input context is the set of free variables in
the term. This has not yet been proven.

@@

\begin{center}
\begin{tabular}{c}
	\infer[\textsc{ceq}]
	{\gamma = \gamma'', \overline{x_i : \mathbb{T}_i} \cand \gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}}
	{\gamma = \gamma' \cand \gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}}
\\\\
	\infer[\textsc{crem}]
	{
	\gamma = \gamma'', \overline{x_i : \mathbb{T}_i} \cand \gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}, x : \mathbb{T}
	}
	{
	\gamma = \gamma' - \{x\} \cand \gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}, x : \mathbb{T}
	}
\\\\
	\infer[\textsc{cext}]
	{\gamma = \gamma'', \overline{x_i : \mathbb{T}_i}, \overline{y_j : \mathbb{T}_j} \cand \gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}}
	{
		\gamma = \gamma', \overline{y_j : \mathbb{T}_j} \cand \gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}
		& \overline{y_j} \cap \overline{x_i} = \emptyset
	}
\\\\
	\infer[\textsc{cext\_2}]
	{\mathbf{bound}\:x}
	{
		\gamma = \gamma', \overline{y_j : \mathbb{T}_j}, x : \mathbb{T} \cand \gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}
		& x \in \overline{x_i}
	}
\\\\
	\infer[\textsc{crep}]
	{\gamma = \gamma'', \overline{x_i : \mathbb{T}_i}, x : \mathbb{T} \cand
	\gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}, x : \mathbb{T}'}
	{\gamma = \gamma'[x \mapsto \mathbb{T}] \cand
	\gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}, x : \mathbb{T}'}
\\\\
	\infer[\textsc{crep\_2}]
	{
	\begin{array}{c}
	\exists \alpha, \gamma_r . 
	\gamma = \gamma'[x \mapsto \mathbb{T}] \cand
	\gamma' = \gamma'', \overline{x_i : \mathbb{T}_i} \cand
	\\
	\gamma'' = \gamma_r, x : \alpha \cand
	\free{\gamma_r}
	\end{array}
	}
	{
	\begin{array}{c}
	\begin{array}{ccc}
	x \notin \overline{x_i}
	&
	\fresh{\gamma_r}
	&
	\fresh{\alpha}
	\end{array}
	\\
	\gamma = \gamma'[x \mapsto \mathbb{T}] \cand
	\gamma' = \gamma'', \overline{x_i : \mathbb{T}_i} \cand
	\free{\gamma''}
	\end{array}
	}
\\\\
	\infer[\textsc{crep\_3}]
	{\mathbf{unbound}\:x}
	{
	\begin{array}{c}
	x \notin \overline{x_i}
	\\
	\gamma = \gamma'[x \mapsto \mathbb{T}] \cand
	\gamma' = \gamma'', \overline{x_i : \mathbb{T}_i} \cand
	\gamma'' = \emptyset
	\end{array}
	}
\\\\
	\infer[\textsc{cvar}]
	{\exists \gamma_r' . 
	\gamma = \gamma_r, \overline{x_i : \mathbb{T}_i} \cand
	\gamma_r = \gamma_r', x : \mathbb{T} \cand
	\free{\gamma_r'}
	}
	{
	\begin{array}{c}
	\begin{array}{cc}
	x \notin \overline{x_i}
	&
	\fresh{\gamma_r'}
	\end{array}
	\\
	\gamma = \gamma_r, \overline{x_i : \mathbb{T}_i} \cand 
	x : \mathbb{T} \in \gamma \cand
	\free{\gamma_r}
	\end{array}
	}
\\\\
	\infer[\textsc{cvar\_2}]
	{\mathbf{unbound}\:x}
	{x : \mathbb{T} \in \gamma \cand \gamma = \gamma', \overline{x_i : \mathbb{T}_i} \cand \gamma' = \emptyset
	&
	x \notin \overline{x_i}}
\end{tabular}
\end{center}
\caption{\label{fig:contextsolve} Context normalisation}
\end{figure}

\begin{figure}
\[
	\infer[\textsc{tvar}]
	{\mathbb{T} = \mathbb{T}' \cand \gamma = \gamma', \overline{x_i : \mathbb{T}_i}, x : \mathbb{T}'}
	{x : \mathbb{T} \in \gamma \cand \gamma = \gamma', \overline{x_i : \mathbb{T}_i}, x : \mathbb{T}'}
\]
\caption{\label{fig:tvargen} Refining type variable constraints}
\end{figure}

>>>>

\newcommand{\typerule}[4]{#1 \triangleright #2 : #3 \triangleleft #4}
\newcommand{\oprule}[4]{#1 \mid #2\;\longrightarrow\;#3 \mid #4}
\newcommand{\inferrule}[4]{\left\llbracket #1 \triangleright #2 : #3 \triangleleft #4 \right\rrbracket}
\newcommand{\inferlhs}[1]{\left\llbracket \gamma \triangleright #1 : \alpha \triangleleft \gamma' \right\rrbracket}
\newcommand{\subst}[3]{#3 \{\overline{^{#1}/_{#2}}\}}
\newcommand{\substo}[3]{#3 \{^{#1}/_{#2}\}}
\newcommand{\substxy}[1]{\subst{x_i}{y_i}{#1}}
\newcommand{\unitv}{\mathtt{unit}}
\newcommand{\unitt}{\mathbf{Unit}}
\newcommand{\truev}{\mathtt{true}}
\newcommand{\falsev}{\mathtt{false}}
\newcommand{\boolt}{\mathbf{Bool}}
\newcommand{\ot}[2]{#1@\overline{#2}}
\newcommand{\funv}[4]{\lambda(\overline{#1 : #2 \gg #3}).#4}
\newcommand{\funt}[3]{(\overline{#1 \gg #2}) \rightarrow #3}
\newcommand{\lett}[3]{\mathtt{let}\:#1\:\mathtt{=}\:#2\:\mathtt{in}\:#3}
\newcommand{\ift}[3]{\mathtt{if} \; #1 \; \mathtt{then} \; #2 \; \mathtt{else} \; #3}

\newcommand{\cand}{\:\wedge\:}
\newcommand{\free}[1]{\mathbf{free}\:#1}
\newcommand{\fresh}[1]{\mathbf{fresh}\:#1}
\newcommand{\partialo}[2]{\mathbf{partial}\:\overline{\mathbb{S}_{#1}{#2}}}
\newcommand{\fixedo}[2]{\mathbf{fixed}\:\overline{\mathbb{S}_{#1}{#2}}}
\newcommand{\os}[2]{\omega_{#1}@\sigma_{#2}}

\newcommand{\tinf}{\mathbb{T}}

@@

% unit value
\inferlhs{\unitv} \triangleq \alpha = \unitt \cand \gamma' = \gamma
\\\\

% true value
\inferlhs{\truev} \triangleq \alpha = \boolt \cand \gamma' = \gamma 
\\\\

% false value
\inferlhs{\falsev} \triangleq \alpha = \boolt \cand \gamma' = \gamma 
\\\\

@@

% let binding
\inferlhs{\lett{x}{t'}{t}} \triangleq \\
\exists \gamma_v, \gamma_{in}, \gamma_{out}, \alpha' . \\
\gamma_{in} = \gamma_v, x : \alpha'
\cand
\gamma' = \gamma_{out} - \{ x \}
\cand \\
\inferrule{\gamma}{t'}{\alpha'}{\gamma_v}
\cand
\inferrule{\gamma_{in}}{t}{\alpha}{\gamma_{out}}
\\\\

@@

Additionally, we have {\bf variable usage constraints} of form 
$x : \tinf \in \gamma$, which are generated for each location in the
program where a variable is used and what type that usage implies. These
are used to generate additional type variable
constraints by matching the asserted type $\tinf$ with what is known
about the type of $x$ in $\gamma$.

@@

Type variables can appear at any position within a type's structure, as shown
in the grammar for $\tinf$. Objects are treated as a special case, where
we use variables as placeholders for the object definition and state parts.
The states and transitions of the object structure are determined at a later
stage in the type inference process.

@@

\begin{itemize}
\item All value terms produce a $\gamma = \gamma'$ constraint as values do
not change the context. Additionally, a type variable constraint of form
$\alpha = \tinf$ where $\tinf$ is of the appropriate type to match
the value, with type variables in all positions relating to parameters and return
values.
\item The context variable $\gamma_a$ used as the input context for
function bodies has an empty root $\gamma_r$, to allow for detection of 
used unbound variables within
the body. This matches the use of a context in type inference
composed only of the parameters, and prevents implicit binding.
\item Let-bind terms generates a new context variable $\gamma_v$ for the
output context of the value term. This is extended by $\gamma_{b1}$, the
input context for the body term, with variable $x$ of the type inferred
for the value. The output context of the whole let-bind term removes
$x$ from the output context $\gamma_{b2}$ of the body term.
\item Update terms generate replacement constraints based on the inferred
type $\alpha'$ of the right-hand side of the expression.
\item Sequencing terms relate the output of the left term to the input of the
right term through a new context variable $\gamma''$. The type variable
$\alpha'$ must be generated to capture the type of the left term for
consistency, but it is not needed.
\item Function call terms generate context variable constraints to relate the
usage of the variables to whatever information has been inferred about the
parameter types of the function. All variables $\overline{x_i}$ used as 
parameters in the function call are replaced in the output context with
new type bindings to $\overline{\alpha_i'}$, which represents the effect
of the function.
\item Method call terms generate type variables to capture that $x$ must
be an object, that the method $m$ must exist in the current state $\alpha_s$
and that after the method call $x$ must be in the new state $\alpha_s'$.
\end{itemize}

\subsection{Constraint solving}

Constraint solving is performed in phases. First, generated context constraints
for a term $t$ aresolved, which determines the contents of each context associated to
each context variable in the typing derivation of $t$, and produces additional
type equality constraints (of form $\mathbb{T} = \mathbb{T}'$) based on
observed usages of variables. If a variable either used in a context where
it is unbound (the variable is {\it free} in $t$) or it is rebound
(a subterm within $t$ is of form $let x = ... in t'$ where a subterm of $t'$
is also of form $let x = ... in t''$), then no principal typing exists and
the phase reports the offending variable.

Then, the set of type constraints is solved, which determines a lower and
upper bound for each type variable, if such bounds can be determined.
Additionally, if the constraints lead to a contradiction (i.e. demanding the
existence of an invalid type like $\boolt \sqcup \unitt$) then no solution
will be reported.

@@

Solving context constraints involves {\it normalising} the constraints. 
Normalisation means that every context variable will have a single 
defining constraint of form $\gamma = \emptyset$, $\free{\gamma}$ or
$\gamma = \gamma', \overline{x_i : \tinf_i}$, where $\gamma'$ is
either empty or free. A context $\Gamma$ for type checking can be easily
constructed from constraints in this form, as once the type variable
constraints are solved we can say $\Gamma = \overline{x_i : T_i}$ for
every $\gamma = \gamma', \overline{x_i : \tinf_i}$ where $T_i$ is the
solved form of $\tinf_i$.

We say that a constraint $C$ is {\it normalisable} if, by
applying the rules in \figref{fig:contextsolve}, it can be transformed to
a constraint $C'$ which is normalised or the rules produce $\mathbf{bound}\:x$
or $\mathbf{unbound}\:x$ for some variable $x$.

@@

\begin{itemize}
\item \textsc{ceq} --- Where $\gamma$ is directly equivalent
to $\gamma'$, which is defined in terms of extension of
$\gamma''$, rewrite the constraints such that
$\gamma$ directly copies the extension of $\gamma''$.
\item \textsc{crem} --- Where $\gamma$ removes a variable $x$ from $\gamma'$,
which is defined in terms of extension of $\gamma''$, rewrite the
constraints such that $\gamma$ copies the extension of $\gamma''$ with
the variable $x$ removed from the extension. There is no need for an
error-producing variant of \textsc{crem}, where the variable $x$ is missing, 
as context removal constraints are always generated in a pair with a context 
extension constraint for let binding.
\item \textsc{cext} --- Where $\gamma$ extends $\gamma'$ with
a variable set $\overline{y_j}$, and $\gamma'$ is defined in terms of extension of
$\gamma''$ with a disjoint set of variables in its extension set, rewrite the
constraints such that $\gamma$'s extension set includes both the $y_j$
definitions and $x_j$ definitions.
\item $\textsc{cext}_2$ --- Where $\gamma$ extends $\gamma'$ with
a new variable $x$, and $\gamma'$ is defined in terms of extension of
some $\gamma''$ and the extension set already contains $x$, this
represents an attempt to redeclare $x$. As such, generate an error
$\mathbf{bound}\:x$, terminating the constraint solving process.
\item \textsc{crep} --- where $\gamma$ replaces the type of some
variable $x$ in $\gamma'$, where $\gamma$ is defined in terms of extension
of $\gamma''$ and $x$ is in the extension set, rewrite the constraints
such that $\gamma$ copies the extension set of $\gamma'$ with $x$
replaced as specified.
\item $\textsc{crep}_2$ --- where $\gamma$ replaces variable $x$ in
$\gamma'$ which is defined in terms of extension of $\gamma''$
which is free and $x$ is not in the extension set, $x$ is a
free variable in the term being processed. To introduce $x$ in such a way that
it will be included in all contexts, $\gamma''$ is redefined in
terms of extension of a new context $\gamma_r$ with variable $x$
in the extension set mapped to a new type variable $\alpha$. Following
this, \textsc{cext} can be applied to $\gamma'$, then \textsc{crep} can
be applied to $\gamma$.
\item $\textsc{crep}_3$ --- where $\gamma$ replaces variable $x$ in
$\gamma'$ which is defined in terms of an empty root $\gamma''$,
and $x$ is not in $\gamma$'s extension set, this indicates the usage
of an unbound variable within a function body. As such, generate an
error $\mathbf{unbound}\:x$ indicating that x is unbound within this
context, terminating the contraint solving process.
\item \textsc{cvar} --- where a variable constraint for $x$ exists
for context variable $\gamma$, which is defined in terms of
extension of $\gamma_r$ which is free, and $x$ is not in the extension
set of $\gamma$, $x$ is a free variable in the term being processed. 
Similar to $\textsc{crep}_2$, redefine $\gamma_r$
in terms of extension of a new root $\gamma_r'$, with $x$ in
the extension set with the observed type $\tinf$.
\item $\textsc{cvar}_2$ --- where a variable constraint for $x$ exists
for context variable $\gamma$, which is defined in terms of $\gamma'$
which is empty, and $x$ is not in the extension set of $\gamma$, 
$x$ is an unbound variable within a function body.
\end{itemize}

@@

@@

\begin{center}
\begin{tabular}{c}
	\infer[\textsc{c\_eq}]
	{\gamma = \gamma'', \overline{x_i : \tinf_i} \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}}
	{\gamma = \gamma' \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}}
\\\\
	\infer[\textsc{c\_eq\_2}]
	{\gamma = \gamma', \emptyset}
	{C \cand \gamma = \gamma' & \free{\gamma'} \in C \vee \gamma' = \emptyset \in C}
\\\\
	\infer[\textsc{c\_rem}]
	{
	\gamma = \gamma'', \overline{x_i : \tinf_i} \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}, x : \tinf
	}
	{
	\gamma = \gamma' - \{x\} \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}, x : \tinf
	}
\\\\
	\infer[\textsc{c\_ext}]
	{\gamma = \gamma'', \overline{x_i : \tinf_i}, \overline{y_j : \tinf_j} \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}}
	{
		\gamma = \gamma', \overline{y_j : \tinf_j} \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}
		& \overline{y_j} \cap \overline{x_i} = \emptyset
	}
\\\\
	\infer[\textsc{c\_ext\_2}]
	{\mathbf{bound}\:x}
	{
		\gamma = \gamma', \overline{y_j : \tinf_j}, x : \tinf \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}
		& x \in \overline{x_i}
	}
\\\\
	\infer[\textsc{c\_rep}]
	{\gamma = \gamma'', \overline{x_i : \tinf_i}, x : \tinf \cand
	\gamma' = \gamma'', \overline{x_i : \tinf_i}, x : \tinf'}
	{\gamma = \gamma'[x \mapsto \tinf] \cand
	\gamma' = \gamma'', \overline{x_i : \tinf_i}, x : \tinf'}
\\\\
	\infer[\textsc{c\_rep\_2}]
	{
	\begin{array}{c}
	\exists \alpha . 
	\gamma = \gamma'[x \mapsto \tinf] \cand
	\gamma' = \gamma'', \overline{x_i : \tinf_i} \cand
	\\
	x : \alpha \in \gamma''
	\end{array}
	}
	{
	\begin{array}{c}
	\begin{array}{cc}
	x \notin \overline{x_i}
	&
	\fresh{\alpha}
	\end{array}
	\\
	\gamma = \gamma'[x \mapsto \tinf] \cand
	\gamma' = \gamma'', \overline{x_i : \tinf_i}
	\end{array}
	}
\\\\
	\infer[\textsc{crep\_3}]
	{C \cand \gamma = \gamma'[x \mapsto \tinf] \cand 
	 x : \alpha \in \gamma'}
	{
	C \cand \gamma = \gamma'[x \mapsto \tinf]
	&
	\free{\gamma'} \in C \vee \gamma' = \emptyset \in C
	}
\\\\
	\infer[\textsc{cvar\_1}]
	{\exists \gamma' . \free{\gamma'} \cand \gamma = \gamma', x : \tinf}
	{\free{\gamma} \cand x : \tinf \in \gamma}
\\\\
	\infer[\textsc{cvar\_2}]
	{\mathbf{unbound}\:x}
	{\gamma = \emptyset \cand x : \tinf \in \gamma}
\\\\
	\infer[\textsc{cvar\_3}]
	{\gamma = \gamma', \overline{x_i : \tinf_i} \cand x : \tinf \in \gamma'}
	{\gamma = \gamma', \overline{x_i : \tinf_i} \cand x : \tinf \in \gamma
	&
	x \notin \overline{x_i}
	}
\\\\
	\infer[\textsc{c\_join}]
	{
	\begin{array}{c}
	\gamma = \overline{x_i : \tinf_i''} \cand 
	\gamma' = \gamma_r, \overline{x_i : \tinf_i} \cand 
	\\
	\gamma'' = \gamma_r, \overline{x_i : \tinf_i'} \cand
	\overline{\tinf_i'' = \tinf_i \sqcup \tinf_i'}
	\end{array}
	}
	{
	\begin{array}{c}
	\gamma = \gamma' \sqcup \gamma'' \cand 
	\gamma' = \gamma_r, \overline{x_i : \tinf_i} \cand 
	\\
	\gamma'' = \gamma_r, \overline{x_i : \tinf_i'}
	\end{array}
	}
\\\\
	\infer[\textsc{cerr\_1}]
	{\mathbf{unbound}\:x}
	{C \cand \mathbf{unbound}\:x}
\\\\
	\infer[\textsc{cerr\_2}]
	{\mathbf{bound}\:x}
	{C \cand \mathbf{bound}\:x}
\end{tabular}
\end{center}
\caption{\label{fig:contextsolve} Context normalisation}
\end{figure}

\subsection{Additional type variable constraint solving}

In the second phase, additional type variable 
constraints are generated by following the rule in \figref{fig:tvargen}, which
compares the observed type from the usage of a variable to the type
known in the context. Repeated application of this rule until no more matches
are found produces a constraint set composed exclusively of normalised
context variable constraints, and type variable constraints. At this point, 
we can focus on the type variable constraints exclusively as the context
variable constraints have no further contribution to the overall solution.

\begin{figure}
\[
	\infer[\textsc{tvar}]
	{\tinf = \tinf' \cand \gamma = \gamma', \overline{x_i : \tinf_i}, x : \tinf'}
	{x : \tinf \in \gamma \cand \gamma = \gamma', \overline{x_i : \tinf_i}, x : \tinf'}
\]
\caption{\label{fig:tvargen} Refining type variable constraints}
\end{figure}

\subsection{Type variable constraint solving}

\subsection{Object variable constraint solving}

Normalised object variable constraints are of the forms listed in
\figref{fig:objconstraints}. Given a set of such normalised
constraints $C_\omega$, solving proceeds by application of the rules in
\figref{fig:objsolving}. Where there is no
explicit partial or fixed solution for a given object variable $\omega$ in
a constraint set, the solution for the object is equivalent to an empty partial solution, $\mathbf{partial}\:\emptyset$.

\begin{figure}
\[
\begin{array}{ll}
\omega = \partialo{}{} 
	& \text{partial object solution}
	\\
\omega = \fixedo{}{}
	& \text{fixed object solution} 
	\\
\sigma = \mathbf{unknown}
	& \text{undefined state set}
	\\
\sigma = \overline{S} 
	& \text{state set solution} 
	\\
\os{1}{1} = \os{2}{2} 
	& \text{equality} 
	\\
\os{1}{1} <: \os{2}{2} 
	& \text{subtyping} 
	\\
\os{1}{1} = \\
\:\:\remapapp{\os{1}{2}}{\os{2}{3}}{\os{2}{4}} 
	& \text{remap definition}
	\\
\os{1}{1} = \os{2}{2} \sqcup \os{3}{3}
	& \text{outer join}
	\\
\os{1}{1} = \os{1}{2} \veebar \os{1}{3}
	& \text{inner join}
\end{array}
\]
\caption{\label{fig:objconstraints} Normalised object variable constraints}
\end{figure}

\begin{figure*}

\begin{center}
\begin{tabular}{cc}

\infer[\textsc{$=$-solve}]
{
	\substo{\omega}{\omega_1}{
	\substo{\omega}{\omega_2}{
	\substo{\sigma}{\sigma_1}{
	\substo{\sigma}{\sigma_2}{C}}}}, \,
	\omega = \Omega, \,
	\sigma = \Sigma
}
{
	\begin{array}{c}
	C, \, 
	\os{1}{1} = \os{2}{2}, \,
	\omega_1 = \Omega_1, \,
	\omega_2 = \Omega_2, \,
	\sigma_1 = \Sigma_1, \,
	\sigma_2 = \Sigma_2
	\\
	\sigma = min(\sigma_1, \sigma_2) 
	\\
	\omega = min(\omega_1, \omega_2) 
	\\
	(\Omega, \Sigma) = \mathbf{connect}(\Omega_1, \Sigma_1, \Omega_2, \Sigma_2)
	\end{array}
}

\\
\\

\infer[\textsc{remap-solve}]
{
	C, \,
	\omega_1 = \partialo{}{'}, \,
	\sigma_1 = \overline{S}_{out}, \,
	\sigma_2 = \overline{S}_{in}
}
{
	\begin{array}{c}
	C, \,
	\os{1}{1} = \remapapp{\os{1}{2}}{\os{2}{3}}{\os{2}{4}}, \,
	\omega_1 = \partialo{1}{}, \,
	\sigma_1 = \Sigma_1, \,
	\sigma_2 = \Sigma_2
	\\
	\omega_2 = \fixedo{2}{} \in C
	\\
	\sigma_3 = \overline{S}_3 \in C
	\\
	\sigma_4 = \overline{S}_4 \in C
	\\
	(\mathbb{S}, \overline{S}_{in}, E) = \mathbf{refine}(\mathbb{S}_1, \Sigma_2, \mathbb{S}_2, \overline{S_3})
	\\
	(\mathbb{S}', \Sigma') = \mathbf{connect}(\partialo{}{}, E(\overline{S_4}), \partialo{}{}, \overline{S}_2)
	\end{array}
}

\\
\\

\infer[\textsc{$<:$-solve}]
{
	C, \,
	\omega_1 = \mathbb{S}', \,
	\sigma_1 = \overline{S}' \,
}
{
	\begin{array}{c}
	C, \,
	\os{1}{1} <: \os{2}{2}, \,
	\omega_1 = \partialo{1}{}, \,
	\sigma_1 = \Sigma_1
	\\
	\omega_2 = \fixedo{2}{} \in C
	\\
	\sigma_2 = \overline{S}_2 \in C
	\\
	(\mathbb{S}', \overline{S}', E) = \mathbf{overlay}(\mathbb{S}_1, \Sigma_1, \mathbb{S}_2, \overline{S}_2)
	\end{array}
}

\\
\\

\infer[\textsc{$\sqcup$-solve}]
{
	C, \,
	\omega_1 = \Omega',
	\sigma_1 = \Sigma'
}
{
	\begin{array}{c}
	C, \,
	\os{1}{1} = \os{2}{2} \sqcup \os{3}{3} \,
	\omega_1 = \Omega_1,
	\sigma_1 = \Sigma_1
	\\
	\omega_2 = \fixedo{2} \in C
	\\
	\omega_3 = \fixedo{3} \in C
	\\
	\sigma_2 = \overline{S}_2 \in C
	\\
	\sigma_3 = \overline{S}_3 \in C
	\\
	(\mathbb{S}_\cup, \overline{S}'_\cup) = \mathbf{union}(\mathbb{S}_2, \overline{S}_2, \mathbb{S}_3, \overline{S}_3)
	\\
	(\Omega', \Sigma') = \mathbf{connect}(\Omega_1, \Sigma_1, \partialo{\cup}{}, \overline{S}_\cup)
	\end{array}	
}

\\
\\
